---
title: "The ecological cost of social progress"
subtitle: "HarvardX Data Science Capstone Own Project"
author: "Codrin Kruijne"
date: "`r Sys.Date()`"
output:
  pdf_document: default
urlcolor: blue
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
require(dplyr)
require(gridExtra)

# Script settings
knitr::opts_chunk$set(
	message = FALSE,
	warning = FALSE,
	echo = FALSE,
	cache = TRUE,
	tidy.opts = list(width.cutoff = 100),
	tidy = TRUE
)
```

This report has been prepared for the [HarvardX Data Science program](https://www.edx.org/professional-certificate/harvardx-data-science) Capstone course final project submission and consists of three files of original work that can be found [on GitHub](https://github.com/codrin-kruijne/HappyPlanetIndex) (with all commits): this [report in PDF](https://github.com/codrin-kruijne/HappyPlanetIndex/blob/master/HarvardXDataScienceCapstoneCodrinKruijneOwnProjectReport.pdf), this [report in R markdown](https://github.com/codrin-kruijne/HappyPlanetIndex/blob/master/HarvardXDataScienceCapstoneCodrinKruijneOwnProjectReport.Rmd) and an [R markdown script file](https://github.com/codrin-kruijne/HappyPlanetIndex/blob/master/HarvardXDataScienceCapstoneCodrinKruijneOwnProjectResults.Rmd) with all the code (also [knitted to PDF with results](https://github.com/codrin-kruijne/HappyPlanetIndex/blob/master/HarvardXDataScienceCapstoneCodrinKruijneOwnProjectResults.pdf) for convenience). Results are read in from data objects stored locally by the script (also saved in the GitHub repository), that were the output of the script file. We were limited in our visualisations by the requirement to publish to PDF; interactive visualisations by publishing to HLTM would have been more insightful.


# Summary

Humankind uses diverse progress indicators, most currently Growth Domestic Product (GDP) to guage economic growth. It is becoming more apparant that this is a limited and myopic view and alternative measures have been emerging over te last years, like the United Nation's Human Development Index. I wanted to have a closer look at two concepts that follow; the Social Progress Index (SPI) and the Happy Planet index (HPI), which take into account the state of our ecosystems. This report describes merging the last avaialbele data for both from the year 2016, applying clustering to see how certain countries group, and having a look at their differences. Resulting clusters show that social progress seems to come at the cost of ecological systems that cannot sustain levels of accompanied consumption.

# Analysis

## Raw data

Data was downloaded from the official websites as Excel files and the last common year, 2016 was read in. HPI data in short looks at the amount of happy life years peole in a country achive, divided by the ecological footprint or 'ecosystem costs' to obtain that to balance means and ends. SPI data includes more diverse indicators with the main categories being Basic Human needs, like health and safety, Foundations for Wellbeing, like access to knowledge and quality environment and, last, Opportuniy, with freedom, rights and inclusiveness. Links to the websites can be foundin references.

## Data preparation

Data from HPI and SPI were joined by Country name. Missing values imputed by mean values. All numeric variables were scaled in preparation for clustering. This leaves us with 118 observations of 87 variables. We explore the data next.

## Data exploration

To get an initial view of the data we plot histograms of HPI scores with different bin withs and a scatter plot with major variables.

```{r Exploratory Histogram}
hpi_hist_3 <- readRDS("hpi_hist_3.rds")
hpi_hist_1 <- readRDS("hpi_hist_1.rds")
hpi_region_hist <- readRDS("hpi_region_hist.rds")
hpi_scatter <- readRDS("hpi_scatter.rds")

grid.arrange(hpi_hist_3, hpi_hist_1)
plot(hpi_region_hist)
```

Looking at all HPI scores the histogram seems to show a somewhat normal distribution. The HPI data is already labeled with regions and separating them clearly shows differences, most notably contrasts like the low scores in Sub Saharan Africa and the distributed, yet high scores of the Americas.

```{r Exploratory scatterplot}
plot(hpi_scatter)
```

The scatterplot shows HPI on the x-axis and the effective footprint on the y-axis. Footprint data represents how big an area of ecosystems is needed (in "Global Hectares per person") to realise the consumption levels in that country (corrected for many things). 1.73 is the theoretical amount our world would be able to sustain. I chose to show the difference on the y-axis; a positive value means countries consume within planetary boundaries; a negative value shows a comsumption rate requires more ecosystem capacity than is available. Here you see differences and the underlying values with notably Sub Saharan africa and Europe at opposite ends. As we have more data from the SPI, I wanted to explore what clustering based on all variables beyond region would yield for insights.

## Clustering modeling

As the data provides six regions, we use that for the amount of initial clusters, after which we will calculate the most appropriate amount of clusters. We used K-means clustering.

```{r Exploratory clusters}
cluster_expl <- readRDS("cluster_expl.rds")
plot(cluster_expl)
```

These clusters calculated on all variables, not just Region, differ quite a bit from the geographical region clusters. Lets calculate the optimal amount of clusters using the Elbow method and Silhouette width.

```{r Optimal Clustering}
elbow_plot <- readRDS("elbow_plot.rds")
sil_width_plot <- readRDS("sil_width_plot.rds")

grid.arrange(elbow_plot, sil_width_plot)
```

Fitting several cluster models (up to k = 10) shows two clusters seems to be the best separation. At k = 2 we see the elbow bend and the average silhouette width is highest when data is separated into two clusters.

# Results

## Visualising clustered data

The figure below show the data with clusters identified by color and size representing first HPI and then SPI scores.

```{r}
bi_plot_hpi <- readRDS("bi_plot_hpi.rds")
plot(bi_plot_hpi)

bi_plot_spi <- readRDS("bi_plot_spi.rds")
plot(bi_plot_spi)
```

The first cluster seems to have most coutries with a positive footprint, meaning their comsumption can be sustained, while the second cluster has increasingly Happy Life Years and more negative footprint. Using HPI as size for datapoints accentuates that the lower happy life years and the higher negative footprint, the less is HPI score, whereas SPI score visually seems similar putting an emphasis on social progress.

## Cluster analysis

To better understand the differences in countries we plot the average scores per cluster on the major components of both the HPI and SPI indicators.

```{r warning=FALSE}
diff_bar <- readRDS("diff_bar.rds")
plot(diff_bar)
```
We can see a clear distinction between clusters on average scores of a select number of indicators. Countries in cluster 1 score above average except for their environmental footprint; they are doing wel socially, but at the cost of our ecological systems. Countries in cluster 2 score well below averages, except for their footprint.

# Conclusions

Although the results here show relative performance of clusters, a major concern is the high number of countries with absolute negative net footprint; meaning we are comsuming not only the services or results of out ecosystems, but the production capacity itself, which will lead to increasing footprint pressure.
Further analysis into all measures may provide additional insights into differences in these clusters. Particularly interesting woud be to explore countries with relatively high social progress and positive net footprint.

# References

1. [How is the Happy Planet Index calculated?](http://happyplanetindex.org/about#how)
2. [Social Progress Imperative](https://www.socialprogress.org/index/global)

# Appendix

```{r}
country_list <- readRDS("country_list.rds")
knitr::kable(country_list)
```