---
title: "Happy Planet Index clustering script"
subtitle: "HarvardX Data Science Capstone Own Project"
author: "Codrin Kruijne"
date: "`r Sys.Date()`"
output:
  pdf_document: default
urlcolor: blue
---

```{r message=FALSE, warning=FALSE, include=FALSE}
require(dplyr)
```

This report has been prepared for the [HarvardX Data Science program](https://www.edx.org/professional-certificate/harvardx-data-science) Capstone course final project submission and consists of three files of original work that can be found [on GitHub](https://github.com/codrin-kruijne/HappyPlanetIndex) (with all commits): this [report in PDF](), this [report in R markdown]() and an [R markdown script file]() with all the code (also [knitted to PDF with results]() for convenience). Results are read in from data objects stored locally (also saved in the GitHub repository), that were the output of the script file.


# Introduction

Humankind uses diverse progress indicators, most currently Growth Domestic Product (GDP) to guage economic growth. It is becoming more apparant that this is a limited and myopic view and alternative measures have been emerging over te last years, like the United Nation's Human Development Index. I wanted to have a closer look at two concepts that follow; the Social Progress Index (SPI) and the Happy Planet index (HPI), which take into account the state of our ecosystems. This report describes merging the last avaialbele data for both from the year 2016, applying clustering to see how certain countries group, and have a look at their differences.

# Analysis

## Raw data
Data was downloaded from the official websites as Excel files and the last common year, 2016 was read in.

## Data preparation
Data from HPI and SPI were joined by Country, scaled and missing values imputed by mean values.

## Data exploration
```{r}

```

## Modeling

```{r}

```


```{r}

```



```{r}

```

# Conclusion


# References
1. [How is the Happy Planet Index calculated?](http://happyplanetindex.org/about#how)
2. [Social Progress Imperative](https://www.socialprogress.org/index/global)