---
title: "Happy Planet Index clustering script"
subtitle: "HarvardX Data Science Capstone Own Project"
author: "Codrin Kruijne"
date: "`r Sys.Date()`"
output:
  pdf_document: default
urlcolor: blue
---

```{r message=FALSE, warning=FALSE, include=FALSE}
require(dplyr)
```

This report has been prepared for the [HarvardX Data Science program](https://www.edx.org/professional-certificate/harvardx-data-science) Capstone course final project submission and consists of three files of original work that can be found [on GitHub](https://github.com/codrin-kruijne/HappyPlanetIndex) (with all commits): this [report in PDF](), this [report in R markdown]() and an [R markdown script file]() with all the code (also [knitted to PDF with results]() for convenience). Results are read in from data objects stored locally (also saved in the GitHub repository), that were the output of the script file.


# Introduction

- HPI
- Datasets
- Clustering
- Results

# Analysis

## Raw data


## Data preparation


## Data exploration
```{r}

```

## Modeling

```{r}

```


```{r}

```



```{r}

```

# Conclusion


# References
1. [How is the Happy Planet Index calculated?](http://happyplanetindex.org/about#how)
2. [Social Progress Imperative](https://www.socialprogress.org/index/global)